---
title: "Reproducible Research: Week 2 Assignment"
author: "Acheampong Owusu"
date: "December 22, 2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducible Research

Setting the working directory to the appropriate folder on my Laptop and loading the relevant libraries.

```{r}
setwd("C:/Users/AOwusu/Documents/")
library(dplyr)
library(ggplot2)
library(magrittr)


```

#Loading and preprocessing the data
Here we load the data and also transform the data (if necessary) into a format which is suitable for our analysis

```{r}
#loading base data
baseData <- read.csv("activity.csv", header = TRUE)

#Checking how the data looks like
str(baseData)

#Data Head
head(baseData)

# Data Dimensions
dim(baseData)

#Data Summary
summary(baseData)
```

#NA and Date String to Date Class modification

```{r}
# changing date type to date class 
baseData$date <- as.Date(as.character(baseData$date))
# creating NA logical vector
baseDataNA <- is.na(baseData$steps)
# creating clean base for later mean calculations
cleanBase <- baseData[!baseDataNA,]

```

#Calculating the mean total number of steps taken per day!

##1. Calculate the total number of steps taken per day

```{r}
# aggregating clean non NA steps per day (SUM)
SummedDataByDay <- aggregate(baseData$steps, by=list(baseData$date), sum)
# adjusting column names
names(SummedDataByDay)[1] ="date"
names(SummedDataByDay)[2] ="totalsteps"
# top 15 of Summed Steps by day
head(SummedDataByDay,15)

```

##2. Making a histogram of the total number of steps taken each day

```{r}
# Plot using ggplot
ggplot(SummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")

```

##3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
# Mean of steps taken per day
mean(SummedDataByDay$totalsteps,na.rm=TRUE)


#median of steps taken per day
median(SummedDataByDay$totalsteps,na.rm=TRUE)

```

# Finding the average daily activity pattern?

## 1. Making a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
nonNASubset <- baseData[!baseDataNA,]
MeanDataByInterval <- aggregate(nonNASubset$steps, by=list(nonNASubset$interval), mean)
# set the column names
names(MeanDataByInterval)[1] ="interval"
names(MeanDataByInterval)[2] ="steps"

ggplot(MeanDataByInterval, aes(x = interval, y=steps)) +
  labs(title = "Sum of Steps by Interval", x = "interval", y = "steps")+
  geom_line(color="red") 

```

## 2. Finding which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval <- MeanDataByInterval[which.max(MeanDataByInterval$steps),]
maxInterval

```

# Inputing missing values
## 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
# generate listing of NA's
missingVals <- sum(baseDataNA)
missingVals

```

## 2. Devising a strategy for filling in all of the missing values in the dataset.

### Strategy: The strategy is to use mean interval steps from Mean Steps for that interval.

## 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
# new dataset
baseData2 <- baseData
# NA's in dataset
# dataset minus NA's for Mean calculation
NABase2 <- baseData2[is.na(baseData2$steps),]
cleanBase2 <- baseData2[!is.na(baseData2$steps),]

# generate Mean Data2 by interval
MeanData2ByInterval <- aggregate(cleanBase2$steps, by=list(cleanBase2$interval), sum)
names(MeanData2ByInterval)[1] ="interval"
names(MeanData2ByInterval)[2] ="steps"

# INPUT METHOD- 2ND ATTEMPT
baseData2 <- baseData
missingData <- is.na(baseData2$steps)
meanVals <- tapply(cleanBase$steps, cleanBase$interval, mean, na.rm=TRUE, simplify=TRUE)
baseData2$steps[missingData] <- meanVals[as.character(baseData2$interval[missingData])]


# original missing data count
sum(missingData)

# count of NA values
sum(is.na(baseData2$steps))

```


## 4a) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
FullSummedDataByDay <- aggregate(baseData2$steps, by=list(baseData2$date), sum)

names(FullSummedDataByDay)[1] ="date"
names(FullSummedDataByDay)[2] ="totalsteps"
head(FullSummedDataByDay,15)

# Plot using ggplot
ggplot(FullSummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
  
# Mean on New Data
mean(FullSummedDataByDay$totalsteps)

# Median on New Data
median(FullSummedDataByDay$totalsteps)

```

## 4b) Do these values differ from the estimates from the first part of the assignment?
### Yes, the mean is the same but the median has risen 1.19 steps.
### Original Mean and Median

```{r}

```


## 4c) What is the impact of imputing missing data on the estimates of the total daily number of steps?
### The effect of using mean data per interval as a data impute method for missing values seems to push overall data towards the mean.

# Are there differences in activity patterns between weekdays and weekends?

```{r}
baseData2$weekday <- weekdays(baseData2$date)
baseData2$weekend <- ifelse (baseData2$weekday == "Saturday" | baseData2$weekday == "Sunday", "Weekend", "Weekday")
#baseData2$weekend <- as.factor(baseData2$weekend)
head(baseData2,5)


MeanDataWeekendWeekday <- aggregate(baseData2$steps, by=list(baseData2$weekend, baseData2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"

ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
 labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
```

### There seems to be variation in the beginning of the day during weekdays, likely due to workplace activities. 

### Also, there seems to be an overall slightly larger incidence of steps during the weekends.
